---
title: "World Population"
author: "Marco Zanardo"
date: "5/8/2022"
output:
  ioslides_presentation:
    css: ../style.css
    incremental: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
# style.css file is the same for the course slides.

knitr::opts_chunk$set(eval = TRUE, include = TRUE, echo = FALSE)
```

## How change the population during time?

The primary goal of this presentation is to understand how the world population change depending on the time of observation and the country. It's interesting to know also the speed of population changing and the amount of people in the future. These goals can be summarized in the following points:

-   world population size;
-   by-country population;
-   population growth rate;
-   future predictions on population.

## Outline

-   world population today;
-   population by state;
-   population in the past;
-   population changing rate;
-   population in the future;
-   sources and citations.

```{r import_tidy, include=FALSE}
# load libraries
library(tidyverse)
library(modelr)
library(broom)
library(ggrepel)
library(gganimate)
library(ggstance)
library(shiny)

# import dataset 
pop <- read_csv("population-since-1800.csv")
regpop <- read_csv("world-population-by-world-regions-post-1820.csv")
pgr <- read_csv("population-growth-rates.csv")
pg <- read_csv("population-growth-the-annual-change-of-the-population.csv")

# tidy data
# country/region population
pop <- pop %>% 
  select(-Code) %>% 
  rename(n = `Population (historical estimates)`)

# only country population
state_pop <- pop %>% 
  filter(Entity != "Africa", Entity != "America", Entity != "South America", Entity != "North America", Entity != "Europe", Entity != "Asia", Entity != "Oceania", Entity != "World") %>% 
  rename(State = Entity)

# only region population
region_pop <- pop %>% 
  filter(Entity == "Africa" || Entity == "America" || Entity == "South America" || Entity == "North America" || Entity == "Europe" || Entity == "Asia" || Entity == "Oceania")

# only World population
world_pop <- pop %>% 
  filter(Entity == "World")

# population since antiquity
past_pop <- regpop %>% 
  select(-Code) %>% 
  rename(n = `Population (historical estimates)`) %>% 
  filter(is.na(n) == FALSE)
```

```{r rate_tidy}
# tidy population rate
# population rate till 2020
pop_rate_perc_2020 <- pgr %>% 
  rename(perc2100 = `Medium fertility variant, 2020 - 2099: Annually interpolated demographic indicators - Population growth rate (percentage)`, perc2020 = `Estimates, 1950 - 2020: Annually interpolated demographic indicators - Population growth rate (percentage)`) %>% 
  select(-Code, -perc2100) %>% 
  filter(Year < 2021)

# population rate from 2021 to 2100
pop_rate_perc_2100 <- pgr %>% 
  rename(perc2020 = `Estimates, 1950 - 2020: Annually interpolated demographic indicators - Population growth rate (percentage)`,
         perc2100 = `Medium fertility variant, 2020 - 2099: Annually interpolated demographic indicators - Population growth rate (percentage)`) %>% 
  select(-Code, -perc2020) %>% 
  filter(Year > 2020)

# account for relevant groups of countries
regions <- c("Africa", "Asia", "Asia, Central", "Caribbean", "Central America", "Central and Southern Asia", "Eastern Asia", "Eastern Europe", "Eastern and South-Eastern Asia", "Europe", "Europe and Northern America", "Europe, Western", "Latin America and the Caribbean", "Middle Africa", "Northern Africa", "Northern Africa and Western Asia", "Northern America", "Northern Europe", "Oceania", "Oceania (excluding Australia and New Zealand)", "Southern Africa", "South America", "South Eastern Asia", "Southern Asia", "Southern Europe", "Sub-Saharan Africa", "Western Asia", "Western Africa", "World")
countries_by_develop <- c("Least Developed Countries", "Small Island Developing States (SIDS)", "More Developed Regions", "Less Developed Regions", "Less Developed Regions, excluding China", "Less Developed Regions, excluding Least Developed Countries", "Land-locked Developing Countries (LLDC)")
countries_by_income <- c("High income countries", "Low-income countries", "Lower-middle-income countries", "Middle-income countries", "No income group available", "Upper-middle-income countries")
groups <- regions %>% append(countries_by_develop) %>% append(countries_by_income)

# tidying pop_rate* into a single data_frame to have a single dataset for rates from 1950 to 2100
pop_rate_perc <- pop_rate_perc_2020 %>% 
  rename(perc = perc2020)
pop_rate_perc2 <- pop_rate_perc_2100 %>% 
  rename(perc = perc2100)
pop_rate <- union(pop_rate_perc, pop_rate_perc2) %>% 
  rename(entity = Entity, year = Year, rate = perc)

```

## How many people are there in the world today?

Here, general information about world population.

```{r world_population}
num_people <- world_pop %>% 
  filter(Year == 2021) %>% 
  select(n)
world_population <- num_people[[1]]
world_population2 <- round(world_population/10^9, digits = 2)
```

Today, the world counts approximately `r world_population` people, about `r world_population2` billions.

```{r wp_plot_present}
world_pop %>% 
  filter(Year > 1950) %>% 
  ggplot(aes(Year, n/10^9)) +
  geom_point(aes(colour = Year, size = n), show.legend = FALSE) +
  labs(title = "World population from 1950 to today",
       subtitle = "Quantities are estimates base on individual country population census",
       caption = "Data from ourworldindata.org",
       x = "Year",
       y = "Population (billions)") +
  theme_minimal()
```

## How changed the number of people in the world in the last two centuries?

* In the last two centuries, population growth had an exponential trend;
* In the first half of 1800 population grew of about 0.5 billions (50 years period of time). On the contrary, the last 50 years saw an increment of about 4 billions, a huge difference (eight times) in the same time interval.

```{r wp_graphic, message=FALSE}
world_pop %>% 
  ggplot(aes(Year, n/10^9)) +
  geom_line() +
  labs(title = "World population from 1800 to today",
       subtitle = "Quantities are estimates base on individual country population census",
       caption = "Data from ourworldindata.org",
       x = "Year",
       y = "Population (billions)") +
  theme_minimal()

```

## What about individual countries?

Let's see how changed the population in individual countries.

```{r countries_pop}
state_pop %>% 
  ggplot(aes(Year, n/10^9)) +
  geom_line(aes(group = State, color = State), alpha = 1/5, show.legend = FALSE) +
  labs(title = "Countries population from 1800 to today",
       subtitle = "Quantities are estimates base on individual country population census",
       caption = "Data from ourworldindata.org",
       x = "Year",
       y = "Population (billions)") +
  theme_minimal()
```

Here we can see there are two countries with more than a billion people today and have a very big margin than all the others countries. These correspond to:

```{r most_populated_states}
most_pop_state <- state_pop %>% 
  filter(!is.na(n)) %>% 
  filter(Year == 2021) %>% 
  arrange(desc(n)) %>% 
  head(2)
```

-   `r most_pop_state[[1]][[1]]` with `r most_pop_state[[3]][[1]]/10^9` and
-   `r most_pop_state[[1]][[2]]` with `r most_pop_state[[3]][[2]]/10^9` billions of people.

## What about specific countries?

Here, let's show how population grew in individual countries like Italy.

```{r country_specific_plot}
poplot <- function(state, year1, year2) {
  state_pop %>% 
  filter(State == state, Year >= year1, Year <= year2) %>%
  ggplot(aes(Year, n/10^6)) +
  geom_line(alpha = 0.5, show.legend = FALSE) +
  scale_size(range = c(2, 20)) +
  scale_x_log10() +
  labs(title = paste(state, year1, "-", year2),
       subtitle = "Quantities are estimates base on individual country population census",
       caption = "Data from ourworldindata.org",
       x = "Year",
       y = "Population (millions)") +
  theme_minimal() 
}

poplot("Italy", 1800, 2020)
```


## What's the trend starting from the antiquity?

```{r world_past_pop}
past_pop %>% 
  filter(Entity=="World") %>% 
  ggplot(aes(Year, n/10^9)) +
  geom_line(aes(group = Entity)) +
  geom_vline(xintercept = 1800, color = "red", size = 1) +
  scale_x_continuous(breaks = seq(-10000, 2500,by = 1000)) +
  labs(title = "World population from the antiquity to today",
       subtitle = "Quantities are estimates base on individual country population census",
       caption = "Data from ourworldindata.org",
       x = "Year",
       y = "Population (billions)") +
  theme_minimal()
```

* This isn't very informative, it only tells that thousands of years ago there were much less people than today and in the last few centuries world population grew a lot!

* The exponential population growth of the last few centuries (i.e. from about 1800) may be caused by two main events:

  * the Renaissance;
  * Industrial Revolution.

## How did population grow in individual countries?

* Let's try to fit a model for a single country like United States. 
* The objective of this and further linear models is to evidence a trend/pattern between year and population size or rate as a predictive model. If a year or period of time, data on population isn't kept then the model can predict a value for this period of time from the data already collected.

```{r us_model}
#{r us_model, out.width = "33%", fig.asp = 1, fig.width = 3, fig.align='default'}
us <- filter(state_pop, State == "United States")
us %>% 
  ggplot(aes(Year, n/10^6)) + 
  geom_line() + 
  ggtitle("Full data") +
  labs(y = "Millions of people") +
  theme_minimal()

us_mod <- lm(n ~ Year, data = us)
us %>% 
  add_predictions(us_mod) %>%
  ggplot(aes(Year, pred/10^6)) + 
  geom_line() + 
  ggtitle("Linear trend") +
  labs(y = "Prediction (mln)") +
  theme_minimal()

us %>% 
  add_residuals(us_mod) %>% 
  ggplot(aes(Year, resid/10^6)) + 
  geom_hline(yintercept = 0, colour = "white", size = 3) + 
  geom_line() + 
  ggtitle("Remaining pattern") +
  labs(y = "Residual (mln)") +
  theme_minimal()
```

## How did population grow in individual countries?

Let's try generate a model for a single country like Italy.

```{r italy_model, out.width = "33%", fig.asp = 1, fig.width = 3, fig.align='default'}
it <- filter(past_pop, Entity == "Italy")
it %>% 
  ggplot(aes(Year, n/10^6)) + 
  geom_line() + 
  ggtitle("Full data") +
  labs(y = "Millions of people") +
  theme_minimal()

it_mod <- lm(n ~ Year, data = it)
it %>% 
  add_predictions(it_mod) %>%
  ggplot(aes(Year, pred/10^6)) + 
  geom_line() + 
  ggtitle("Linear trend") +
  labs(y = "Prediction (mln)") +
  theme_minimal()

it %>% 
  add_residuals(it_mod) %>% 
  ggplot(aes(Year, resid/10^6)) + 
  geom_hline(yintercept = 0, colour = "white", size = 3) + 
  geom_line() + 
  ggtitle("Remaining pattern") +
  labs(y = "Residual (mln)") +
  theme_minimal()
```

From the chart of Italy growth, it can be seen that there's an unusual pattern between year zero and year 1750. Let's focus on that area.

```{r past_specific_italy_model}
it <- filter(past_pop, past_pop$Entity == "Italy", between(Year, 0, 1750))
it %>% 
  ggplot(aes(Year, n/10^6)) + 
  geom_line() + 
  ggtitle("Full data") +
  labs(y = "Population (mln)") +
  theme_minimal()

it_mod <- lm(n ~ Year, data = it)
it %>% 
  add_predictions(it_mod) %>%
  ggplot(aes(Year, pred/10^6)) + 
  geom_line() + 
  ggtitle("Linear trend") +
  labs(y = "Prediction (mln)") +
  theme_minimal()

it %>% 
  add_residuals(it_mod) %>% 
  ggplot(aes(Year, resid/10^6)) + 
  geom_hline(yintercept = 0, colour = "white", size = 3) + 
  geom_line() + 
  geom_vline(colour = "red", size = 2, xintercept = 476) + 
  geom_vline(colour = "red", size = 2, xintercept = 1300) + 
  ggtitle("Remaining pattern") +
  labs(y = "Residual (mln)") +
  theme_minimal()
```

From this chart it's clear that some reasons for this unexpected behavior are:

* the fall of Western Roman Empire in 476 A.D. and
* epidemics and famines in the Middle Ages.

## What's the trend for each country?

Here, the objective is to compute a model for all countries and show which of these best fit and worst fit to the model.

```{r forall_model, warning=FALSE, message=FALSE, error=FALSE}
by_country <- state_pop %>% 
  group_by(State) %>% 
  nest()

# function that apply a linear model with year as independent variable and population size (n) as dependent variable
country_model <- function(df) {
  lm(n ~ Year, data = df)
}

by_country <- by_country %>% 
  mutate(model = map(data, country_model))

# # Show an example
# by_country %>% 
#   filter(State == "United States")

# by_country %>% 
#   arrange(State)

# add residuals
by_country <- by_country %>% 
  mutate(data = map2(data, model, add_residuals))

# unnest statistics and force the dropping of alternative list-columns with .drop
resids = unnest(by_country, data, .drop = TRUE)

resids %>% 
  ggplot(aes(Year, resid/10^6)) +
  geom_line(aes(group = State), alpha = 1 / 3) + 
  theme_minimal() + 
  labs(title = "Residuals for all countries",
       caption = "Data from ourworldindata.org",
       x = "Year",
       y = "Residual (mln)")

```

```{r statistics, warning=FALSE, message=FALSE, error=FALSE}
# add statistics
by_country <- by_country %>% 
  mutate(glance = map(model, broom::glance))

# unnest statistics and force the dropping of alternative list-columns with .drop
glance = unnest(by_country, glance, .drop = TRUE)

# bad_fit <- filter(glance, r.squared < 0.25)
# 
# state_pop %>% 
#   semi_join(bad_fit, by = "State") %>% 
#   ggplot(aes(Year, n, colour = State)) +
#     geom_line() +
#     theme_minimal()


bad_fit <- filter(glance, r.squared < 0.30)

state_pop %>% 
  semi_join(bad_fit, by = "State") %>%
  filter(Year > 1950) %>% 
  ggplot(aes(Year, n/10^6, colour = State)) +
    geom_line() +
    theme_minimal() +
    labs(y = "Population (mln)",
       caption = "Data from ourworldindata.org",
       title = "Countries that worst fit the model")

good_fit <- filter(glance, r.squared > 0.98)

state_pop %>% 
  semi_join(good_fit, by = "State") %>% 
  filter(Year > 1950) %>% 
  ggplot(aes(Year, n/10^6, color = State)) +
  geom_line(aes(group = State)) +
  theme_minimal() +
  labs(y = "Population (mln)",
       caption = "Data from ourworldindata.org",
       title = "Countries that best fit the model")

```

## How fast is the population growing?

Here, the purpose is to analyse the rate of population growing and its relationship with the actual number of people (population).

> Definition (annual population growth rate): The annual average rate of change of population size, for a given country, territory, or geographic area, during a specified period.

The growth rate is the percentage of the ratio between annual population increase and total population size. The increase is the sum of:

* difference between births and deaths;
* difference between immigrants and emigrants.

That is refereed to an entity, like a country, region or the entire world, in a specific period of time.

```{r present_rate}
pop_rate_perc_2020 %>% 
  group_by(Year) %>% 
  summarise(n = mean(perc2020)) %>% 
  ggplot(aes(Year,n)) +
  geom_line() +
  labs(title = "Population growth rate since 1950",
       subtitle = "Data resulting from the mean of various countries rates",
       caption = "Data from ourworldindata.org",
       x = "Year",
       y = "Rate") + 
  theme_minimal()

# today overall rate
value_tb1 <- pop_rate_perc_2020 %>% 
  filter(Year == 2020) %>% 
  summarise(mean(perc2020))

#value_tb1[[1]]

value_tb <- pop_rate_perc_2100 %>% 
  filter(Year == 2022) %>% 
  summarise(mean(perc2100))

#value_tb[[1]]

# pop_rate_perc_2020 %>%
#  group_by(Year) %>%
#  summarise(mean(perc2020)) %>%
#  ggplot(aes(Year, `mean(perc2020)`)) +
#  geom_line()

# Italy's 2020 rate
italy_today_rate <- pop_rate_perc_2020 %>% 
  filter(Entity == "Italy" & Year == 2020)

rate1962 <- pop_rate_perc_2020 %>% 
  filter(Year == 1962) %>% 
  summarise(n = mean(perc2020))
```

* World population reached a peak in 1962 with a growing rate of `r rate1962[[1]]`.

* This year the world population is still growing because the mean growing rate is `r value_tb[[1]]`. 
* Population is growing slowly with respect to past decades and this can be seen from the chart where there's the mean of rate in the last 70 years. The population over the years is growing more and more slowly because the growth rate is decreasing.

## What will happen to world population in the future?

How will change world population in the next years/decades? At what pace population will change? Will world population increase forever?

```{r future_population}
pop_rate_perc_2100 %>% 
  filter(Entity == "World") %>% 
  ggplot(aes(Year,perc2100)) + 
    geom_line() +
  labs(title = "Forecasts on rate of population growth from 2022",
       subtitle = "Data resulting from an overall summary of statistics",
       caption = "Data from ourworldindata.org",
       x = "Year",
       y = "Rate") + 
  theme_minimal()

```

In the next years, world population growth rate will decrease, this means that population will increase in number but slowly than before. It can be seen that predictions estimate a slow decreasing of world growth rate in the next decades. This means that:

1.  population is still increasing in number because the rate is positive;
2.  population will grow more and more slowly because the rate is decreasing;
3.  if the rate will reach zero at about year 2100, then population will remain constant in number;
4.  finally, if the rate will become negative, then population will begin to decrease, inevitably.

## How will the growth rate of population change in the future?

```{r region_rate}
# Rate
pop_rate_perc2 %>% 
  filter(Entity %in% regions) %>% 
  ggplot(aes(Year,perc,color=Entity)) +
  geom_line(show.legend = FALSE) +
  geom_ref_line(1, colour = "red") +
  labs(title = "Growth rate of world regions",
       subtitle = "Data based on United Nations (UN) projections",
       caption = "Data from ourworldindata.org",
       x = "Year",
       y = "Rate") +
  theme_minimal()
```

From the plot it can be seen that before 2100 all regions of the world will have a rate less than 1, this means that *no* region will increase its population faster then before.

```{r income_rate}
# Income
income_country_df <- pop_rate_perc2 %>% 
  filter(Entity %in% countries_by_income, Year == 2099)

pop_rate_perc2 %>% 
  filter(Entity %in% countries_by_income) %>% 
  ggplot(aes(Year,perc,color=Entity)) +
  geom_line() +
  labs(title = "Growth rate of countries based on income",
       subtitle = "The gross national income (GNI), previously known as gross national product (GNP), is the total \ndomestic and foreign output claimed by residents of a country, consisting of gross domestic \nproduct (GDP), plus factor incomes earned by foreign residents, minus income earned in the \ndomestic economy by nonresidents.",
       caption = "Data from ourworldindata.org",
       x = "Year",
       y = "Rate") +
  theme(legend.position = 'bottom')
```

* low-income countries have a relatively high growth rate than high-income countries, in 2020. 
* this difference is going to reduce with time because low-income countries will see their growth rate go down and the one of high-income countries will remain approximately the same.

```{r develop_rate}
# Develop
develop_country_df <- pop_rate_perc2 %>% 
  filter(Entity %in% countries_by_develop, Year == 2030)

pop_rate_perc2 %>% 
  filter(Entity %in% countries_by_develop) %>% 
  ggplot(aes(Year,perc,color=Entity)) +
  geom_line(show.legend = FALSE) +
  geom_label_repel(data = develop_country_df,
                   mapping = aes(label = Entity),
                   show.legend = FALSE,
                   alpha = 0.5) +
  labs(title = "Growth rate of regions based on development",
       subtitle = "Data based on United Nations (UN) projections",
       caption = "Data from ourworldindata.org",
       x = "Year",
       y = "Rate") +
  theme_minimal()
```

* least-developed countries start from a growth rate above two and go down to one half in the next 80 years. 
* more-developed regions will remain with the same growth rate, around zero. 
* note how China alone can reduce significantly the growth rate of less-developed regions. This because China has a particular control over population (e.g. China's family planning policies). An example of such policies is *one-child policy*, an official program initiated in the late 1970s and early '80s by the central government of China, the purpose of which was to limit the great majority of family units in the country to one child each. The rationale for implementing the policy was *to reduce the growth rate* of China's enormous population.

## What is the pattern of world growth rate in the next 80 years?

World population growth depends on two metrics:

-   number of births;
-   number of deaths.

```{r future_rate_model}
# reg_rate_perc <- pop_rate_perc2 %>% 
#   filter(Entity %in% regions)
reg_rate_perc <- pop_rate_perc2 %>% 
  filter(Entity == "World")

reg_rate_mod <- lm(perc ~ Year, data = reg_rate_perc)

reg_rate_perc <- reg_rate_perc %>% add_predictions(reg_rate_mod)
reg_rate_perc <- reg_rate_perc %>% add_residuals(reg_rate_mod)

# Predictions
# reg_rate_perc %>% 
#   ggplot(aes(Year, pred)) +
#   geom_line()
reg_rate_perc %>% 
  ggplot(aes(Year)) +
  geom_line(aes(y = perc), color = "blue") +
  geom_line(aes(y = pred), color = "red") +
  labs(title = "Global growth rate (blue line) and pattern (red line) in the future",
       subtitle = "Data based on United Nations (UN) projections",
       caption = "Data from ourworldindata.org",
       x = "Year",
       y = "Rate") +
  theme_minimal()

# Residual
reg_rate_perc %>% 
  ggplot(aes(Year)) +
  geom_line(aes(y = resid), color = "green") +
  labs(title = "Global growth rate residual",
       subtitle = "Data based on United Nations (UN) projections",
       caption = "Data from ourworldindata.org",
       x = "Year",
       y = "Residual rate") +
  theme_minimal()
  
```

* In the future of population, the rate of population growth will turn negative, this means that there will be no increase in population and on the contrary, population may decrease in number.

* The causes of this deceleration may be searched in 
  * the impact on environment of a large population 
  * the reduced resources and food 
  * the longer life expectancy (deriving from an improved health) 
  * the higher cost of living.

* If in the past the factor that check population growth was the mortality, it seems that in the future will be very *low fertility* that will control population changes.

## Sources and Citations

-   [Our world in data](https://ourworldindata.org/world-population-growth);
-   [United Nations Population Database](http://esa.un.org/unpd/wpp/);
-   Max Roser, Hannah Ritchie and Esteban Ortiz-Ospina (2013) - "World Population Growth". Published online at [OWID](OurWorldInData.org). Retrieved from: [Online Resource](https://ourworldindata.org/world-population-growth);
-   [Wikipedia](https://en.wikipedia.org/wiki/Population_growth#Population_growth_rate);
-   [World Health Organization](https://www.who.int/data/gho/indicator-metadata-registry/imr-details/1120).
-   [Future population growth](https://ourworldindata.org/future-population-growth)















